Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-09-25 paper/introduction.tex
wip

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. To combine the benefits of forward and backward modeling, several authors have proposed to use cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA) \citep{dechevigne2018CCA}.
CCA and PLS aim to find, from two sets of data $X$ and $Y$, the components $v$ and $w$ were $Xw$ and $Yv$ are maximially correlated or maximally covarying respectively. Because CCA and PLS are are based on a generalized eigen decomposition, their resulting coefficients are mixing the features of $X$ and $Y$ in a way that makes them notoriously difficult to interpret \citep{lebart1995statistique}.

Here, we introduce the `back-to-back regression', which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide a fast, robust and interpretable causal coefficients. After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain responses (Section~\ref{sec:experiment_real}).

>>>>

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. To combine the benefits of forward and backward modeling, several authors have proposed to use cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA) \citep{de2019multiway}.
CCA and PLS aim to find, from two sets of data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximially correlated or maximally covarying respectively. Because CCA and PLS are are based on a generalized eigen decomposition, their resulting coefficients are mixing the features of $X$ and $Y$ in a way that makes them notoriously difficult to interpret \citep{lebart1995statistique}.

Here, we introduce the `back-to-back regression' (B2B), which not only combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}), but also provide robust and interpretable causal coefficients. After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques in identifying causal influence (Section~\ref{sec:experiment_synthetic}). Finally, we apply B2B to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain representations (Section~\ref{sec:experiment_real}).

### 2019-09-25 paper/paper.tex
wip

\begin{abstract}
Identifying causes from observations can be particularly challenging when i) potential factors are difficult to manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
method designed to efficiently measure, from a set of co-varying factors, the causal
influences that most plausibly account for multidimensional observations. After
proving the consistency of B2B and its links to other linear approaches, we show that our method outperforms
least-squares regression and cross-decomposition techniques (e.g.
canonical correlation analysis and partial least squares) on causal
identification. Finally, we apply B2B to
neuroimaging recordings of 102 subjects reading word sequences. The results
show that the early and late brain responses caused by low- and high-level
word features respectively are more reliably detected with B2B than with  other standard techniques.
% conclude  open
\end{abstract}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F,\label{eq:model}
\end{equation}
%
where $N$ is a centered, unobserved noise variable.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
% what is a general covariance matrix?
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2B) is to estimate the matrix of causal influences $E$.

>>>>

\begin{abstract}
Identifying causes from observations can be particularly challenging when i) potential factors are difficult to manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
method designed to efficiently measure, from a set of co-varying factors, the causal
influences that most plausibly account for multidimensional observations. After
proving the consistency of B2B and its links to other linear approaches, we show that our method outperforms
least-squares regression and cross-decomposition techniques (e.g.
canonical correlation analysis and partial least squares) on causal
identification. Finally, we apply B2B to
neuroimaging recordings of 102 subjects reading word sequences. The results
show that the early and late brain representations, caused by low- and high-level
word features respectively, are more reliably detected with B2B than with other standard techniques.

\end{abstract}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, the problem can be formalized as:
%
\begin{equation}
    Y = (XE + N)F,\label{eq:model}
\end{equation}
%
where $N$ is a centered, unobserved noise variable.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
% what is a general covariance matrix?
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2B) is to estimate the matrix of causal influences $E$.
